---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Rfmalloc

Rfmalloc is an experimental R package for allocating vector storage from a
memory-mapped backing file using a patched copy of
[fmalloc](https://github.com/yasukata/fmalloc). The goal is to explore large,
file-backed R vector storage without routing large allocations back through the
process heap.

The current package exposes an `Rf_allocVector3()`-based custom allocator for
integer, numeric, and logical vectors. This is useful for experiments and stress
testing, but `Rf_allocVector3()` is not part of R's public API, so this package
should still be treated as research software.

## Current Status

Recent allocator work fixed the stress-path that previously fell back into
ptmalloc/dlmalloc's system `mmap()` path for large requests. Large allocations
are now carved from contiguous runs in the fmalloc backing file instead.

Implemented now:

- file-backed allocation for integer, numeric, and logical vectors;
- large allocations spanning multiple fmalloc chunks;
- `malloc`, `free`, and `realloc` support through the patched fmalloc layer;
- reopening backing files created by the current allocator format;
- explicit rejection of older incompatible fmalloc backing files.

Still experimental / future work:

- ALTREP wrappers so R-level duplication, subsetting, and copy-on-write can stay
  file-backed;
- richer persistence semantics and metadata;
- support for more vector types;
- production-grade lifecycle management for mapped files.

## Installation

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("sounkou-bioinfo/Rfmalloc")
```

## Basic Usage

```{r}
library(Rfmalloc)

alloc_file <- tempfile(fileext = ".bin")
init_fmalloc(alloc_file)

v_int <- create_fmalloc_vector("integer", 10)
v_num <- create_fmalloc_vector("numeric", 10)

v_int[1:3] <- c(1L, 2L, 3L)
v_num[1:3] <- c(1.1, 2.2, 3.3)

v_int[1:3]
v_num[1:3]

rm(v_int, v_num)
gc()
cleanup_fmalloc()
unlink(alloc_file)
```

## Larger Allocation Example

```{r}
library(Rfmalloc)

large_file <- tempfile(fileext = ".bin")
init_fmalloc(large_file, size_gb = 0.1)

# About 20 MB of integer payload, larger than the historical 16 MB chunk limit.
big_int <- create_fmalloc_vector("integer", 5e6)
big_int[1:5] <- 1:5
big_int[1:5]

rm(big_int)
gc()
cleanup_fmalloc()
unlink(large_file)
```

## Why ALTREP Is Still the Likely Next Step

Allocating the initial vector storage in the backing file is not enough to make
all R operations file-backed. R's normal duplication paths often call
`allocVector()`, not `Rf_allocVector3()`, so copy-on-write or subsetting may
produce ordinary heap-backed vectors.

A UFO/Travel-like direction is still interesting, but for this package the next
practical layer is likely ALTREP: implement `Dataptr`, `Elt`, `Get_region`, and
especially `Duplicate` so R can preserve file-backed storage during ordinary
vector operations. Pointer-access instrumentation is only needed if we want more
transparent interception than ALTREP provides.

## References

- fmalloc: <https://github.com/yasukata/fmalloc>
- Simon Urbanek's mmap allocator example: <https://gist.github.com/s-u/6712c97ca74181f5a1a5>
- UFOs: <https://github.com/PRL-PRG/UFOs>
- Travel: <https://github.com/Jiefei-Wang/Travel>

## License

GPL (>= 2)
