---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Rfmalloc: out-of-core arrays for R

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/Rfmalloc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rfmalloc/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

**One repo, one story: array computation on data that does not fit in RAM,
over pluggable storage codecs and compute backends.**

Memory-mapped file storage (a patched [fmalloc](https://github.com/yasukata/fmalloc)
allocator behind R's ALTREP) turns RAM from a hard cutoff into a spectrum of
speed levels. On top of that substrate sit two plugin registries that everything
else closes over:

- a **codec registry** - how bytes are stored: lossless (`f64` `f32` `f16`
  `bf16` `alp` `sparse`) for scientific data where bit-exactness is
  non-negotiable, and lossy GGUF quantized blocks (`q4_0` … `q6_k`) for model
  weights at 2-8 bits each;
- a **matmul backend registry** - how products are computed: R's BLAS by
  default, [GGML](https://github.com/ggml-org/ggml) natively in quantized
  space, GPU backends next.

The contract that makes it compose is the **decline protocol**: any backend may
refuse any product, and the substrate falls back to bounded decode panels +
BLAS. Plugins change how fast you get an answer, never whether it is correct.

## Two domains, one substrate

The same file-backed matrices and the same backend registry serve both
larger-than-RAM **scientific computing** and quantized **LLM inference** - the
whole point of the abstraction. Both examples below are `%*%`/`crossprod` over
fmalloc tensors; the only thing that differs is which codec and backend the
registry dispatches to.

### Genomics: out-of-core PCA (runs at render time)

A cells × genes expression matrix lives in a memory-mapped file, so it scales
past RAM (products stream in panels). Highly-variable-gene selection and PCA
never materialize an ordinary R copy, and the linear algebra goes through the
backend registry:

```{r pca, message = FALSE}
library(Rfmalloc)

rt <- open_fmalloc(tempfile(fileext = ".bin"), size_gb = 1)
set.seed(1)
# 3000 cells x 600 genes with 4 latent "cell-type" factors + noise; file-backed
factors  <- matrix(rnorm(3000 * 4), 3000, 4)          # per-cell factor scores
loadings <- matrix(rnorm(4 * 600), 4, 600)            # per-gene factor loadings
X <- create_fmalloc_matrix("numeric", nrow = 3000, ncol = 600, runtime = rt)
X[] <- 3 * (factors %*% loadings) + matrix(rnorm(3000 * 600), 3000, 600)

vars <- fmalloc_colVars(X)          # per-gene variance, single-pass reduction (no R copy)
length(which(vars > quantile(vars, 0.9)))   # candidate highly-variable genes

pca <- fmalloc_pca(X, k = 6)        # Gram matrix + scores, dispatched through the backend registry
round(pca$sdev, 2)                  # 4 factors stand out above the noise floor
dim(pca$x)                          # cell scores in the top PC space
```

### Someone else's algorithm, our storage (runs at render time)

`bigPCAcpp` and `bigPLSR` link bigmemory; [`pcaone`](https://github.com/Zilong-Li/PCAoneR)
takes an `Eigen::Map<MatrixXd>`. All three consume a bare `double*`. An fmalloc
ALTREP matrix *is* a `double*`, into an mmap. So a randomized SVD written by
somebody who has never heard of this package runs out-of-core over it, unchanged,
with no copy:

```{r pcaone, message = FALSE, eval = requireNamespace("pcaone", quietly = TRUE)}
library(pcaone)                       # not ours: Li et al. 2023, randomized SVD

rt2 <- open_fmalloc(tempfile(fileext = ".bin"), size_gb = 3)
m <- 2e6L; n <- 40L
G <- create_fmalloc_matrix("numeric", nrow = m, ncol = n, runtime = rt2)
set.seed(1); for (j in seq_len(n)) G[, j] <- rnorm(m)   # 610 MB, on disk

invisible(gc(reset = TRUE))
sv_mapped <- pcaone(G, k = 5)$d                         # straight off the mmap
peak_mapped <- gc()["Vcells", "max used"] * 8 / 2^20

invisible(gc(reset = TRUE))
sv_heap <- pcaone(matrix(G[], m, n), k = 5)$d           # an ordinary R matrix
peak_heap <- gc()["Vcells", "max used"] * 8 / 2^20

round(c(payload_MB    = m * n * 8 / 2^20,
        peak_heap_MB  = peak_heap,          # pays for the payload
        peak_mmap_MB  = peak_mapped))       # does not
identical(sv_mapped, sv_heap)               # and the answer is the same one
```

The payload never enters the R heap. Nothing in `pcaone` knows why.

That is the deal for anything already written against a pointer. Where it stops
working is where no pointer can exist: **compressed** data. Nobody can hand
`pcaone` a pointer into a PLINK `.bed`, because a genotype there is two bits, not
eight bytes. That is what the codec registry is for, and it is the only thing it
is for:

```{r bed, message = FALSE}
rt3 <- open_fmalloc(tempfile(fileext = ".bin"), size_gb = 1)
set.seed(2)
# 5000 samples x 1000 variants of hardcalls (0/1/2), SNP-major like a real .bed
g  <- matrix(sample(c(0L, 1L, 2L), 5e6, replace = TRUE), 5000, 1000)
tn <- fmalloc_bed(g, runtime = rt3)          # 2 bits per genotype, memory-mapped

c(bits_per_genotype = 8 * length(unclass(tn)) / length(g),
  vs_double         = length(g) * 8 / length(unclass(tn)))

# Products decode bounded column panels and hand them to BLAS. The genotypes are
# never materialized as doubles; these two kernels are what a randomized SVD needs.
Omega <- matrix(rnorm(1000 * 5), 1000, 5)
max(abs(matrix((tn %*% Omega)[], 5000, 5) - matrix(as.numeric(g), 5000, 1000) %*% Omega))
```

```{r cleanup_pcaone, include = FALSE, eval = requireNamespace("pcaone", quietly = TRUE)}
cleanup_fmalloc(rt2)
```

```{r cleanup_bed, include = FALSE}
cleanup_fmalloc(rt3)
```

That codec is one of a family, and the family divides along a real line. `fmalloc_bed`
stores PLINK hardcalls at 2 bits, `fmalloc_dosage` stores fractional dosages at
1 byte, and both **fuse** per-variant centering and scaling into the decode: one
lookup-table read per genotype does decode, mean-imputation (a missing call maps
to the variant mean, hence to 0 after centering, so imputation is free) and
standardization at once. `fmalloc_bed_standardize()` therefore turns a genotype
panel into a PCA-ready standardized matrix with no genotype ever expanded to a
double. That is the **numeric** side, and it is what the matmul codec ABI serves.

Phase and multiallelic data are not numbers, so they get a different interface.
`fmalloc_haplotypes` stores phased haplotypes at 1 bit each and is deliberately
**not** a matmul codec: it is a sibling over the same file-backed storage, with
typed accessors instead of decode-to-double, and it feeds haplotype HMMs like
[`kalis`](https://kalis.louisaslett.com/) (Li and Stephens local ancestry)
losslessly, the way `pcaone` consumes the numeric side. Two access interfaces,
one storage substrate. And `Rpgen` brings a native PLINK2 `.pgen` reader in C
(it vendors pgenlib, GPL-3), so both interfaces can be fed from the on-disk
format that carries genotypes, dosages, phase and multiple alleles at once.

### LLM: bytes in, bytes out

A GGUF model's weights are memory-mapped and stay quantized; the transformer
runs through GGML's SIMD kernels with a KV cache, and the I/O boundary is raw
bytes (the tokenizer is an edge codec built from GGUF metadata - text is the
caller's interpretation, `rawToChar()`):

```{r llm, eval = FALSE}
library(Rllm)

rt <- Rfmalloc::open_fmalloc("rllm-work.bin", size_gb = 2)
model <- rllm_gguf_model("SmolLM2-135M.Q4_K_M.gguf", runtime = rt)
# Every weight borrows its exact read-only span in the original GGUF mapping.
# The encoded q4_k/q5_0/q6_k/q8_0 bytes are neither decoded nor copied into a
# second backing file, and the forward pass points GGML tensors at them directly.

gen <- rllm_generate(model, charToRaw(
    "The capital of France is Paris. The capital of Germany is"), n_new = 16L,
    runtime = rt) # the optional KV cache, not the weights, uses this store
rawToChar(gen$raw)
#> [1] " Berlin. The capital of Italy is Rome. The capital of Spain is Madrid."
```

That is a real recorded run (SmolLM2-135M `Q4_K_M`, 30 layers, grouped-query
attention): ~16 tokens/s CPU decode with the KV cache, 8.5× over cache-less
re-forwards, and - checked against a pure-R reference forward - numerically
exact to the quantization.

## Compute backends: today and next

Because both PCA and inference dispatch their products through the **one**
backend registry, a faster backend accelerates both at once, with no change to
`fmalloc_pca()` or `rllm_generate()`:

- **today, CPU** - R's BLAS for dense products; GGML's runtime-SIMD-dispatched
  quantized kernels (AVX2 on x86, NEON on aarch64, CPUID-selected - no
  non-portable flags) for quantized weights; both out-of-core.
- **today, GPU: Vulkan.** GGML's Vulkan backend is vendored and builds into
  `Rggml` **on request** - `install.packages("Rggml", configure.args =
  "--with-vulkan")`, needing `glslc` and the Vulkan headers. It is not
  auto-detected, because GGML embeds 156 compiled SPIR-V shaders and the
  largest is a 141 MB array literal costing ~5 GB of RAM to compile. It runs on
  any vendor's GPU (NVIDIA, AMD, Intel). `rggml_vulkan_info()` reports what it
  sees; a build without it reports zero devices rather than failing, so callers
  can probe and fall back. Correctness is pinned against the CPU backend, and
  can be exercised without a GPU at all through Mesa's software driver
  (`GGML_VK_ALLOW_CPU=1`).
- **next: CUDA.** NVIDIA-only, and the fastest: the same SmolLM2 `Q4_K_M`
  decodes at **~455 tok/s (~28× this CPU stack)** on an RTX 5050 via upstream
  GGML CUDA. The device-buffer residency API the Vulkan backend needed
  (`Rggml_backend_alloc_ctx_tensors`/`tensor_set`/`tensor_get`) is backend
  agnostic, so CUDA reuses it; what remains is vendoring `ggml-cuda` at a
  version matching the core and adding `nvcc` rules to the build.

## The packages

Six packages, one stack. Each links to its reference documentation; the
sources live under [`packages/`](https://github.com/sounkou-bioinfo/Rfmalloc/tree/main/packages)
in this repository.

| package | role |
|---|---|
| [**Rfmalloc**](https://sounkou-bioinfo.github.io/Rfmalloc/Rfmalloc/) | the substrate: file-backed ALTREP vectors/matrices/tensors, codec + backend registries, panel matmul, out-of-core eviction, `fmalloc_pca()`, and the genotype codecs (`bed`/`dosage`/`haplotypes`) with fused standardization |
| [**Rgguf**](https://sounkou-bioinfo.github.io/Rfmalloc/Rgguf/) | GGUF model files through GGML's official parser: metadata, tensor directory, explicit decoded copies, encoded copies, or borrowed read-only tensor views |
| [**Rggml**](https://sounkou-bioinfo.github.io/Rfmalloc/Rggml/) | compute and format authority: vendored GGML kernels and official GGUF implementation, with BLAS and runtime-SIMD-dispatched quantized backends |
| [**Rllm**](https://sounkou-bioinfo.github.io/Rfmalloc/Rllm/) | composition + inference: registers Rggml as Rfmalloc's codec-aware matmul backend, and builds the llama forward pass, KV cache, and byte-level generation |
| [**Rpgen**](https://github.com/sounkou-bioinfo/Rfmalloc/tree/main/packages/Rpgen) | genomics I/O: bounded ingestion from PGEN, BED, PED/MAP, TPED/TFAM, BGEN, VCF/BCF, GEN, HAPS/legend, EIGENSTRAT, and legacy dosage into hardcall, dosage, haplotype, or dense stores |
| [**RfmallocStatgen**](https://github.com/sounkou-bioinfo/Rfmalloc/tree/main/packages/RfmallocStatgen) | statistical genetics over those stores: streamed regression/PCA, banded LD, LDpred2, and matrix-form colocalisation |

## Install

From the [r-universe](https://sounkou-bioinfo.r-universe.dev):

```{r install, eval = FALSE}
install.packages("Rllm",
  repos = c("https://sounkou-bioinfo.r-universe.dev", getOption("repos")))
```

or a GitHub subdir ref: `pak::pak("sounkou-bioinfo/Rfmalloc/packages/Rllm")`.

All six packages target Linux, macOS and Windows (Rtools), and all six are
verified on ARM in CI: Apple Silicon and Linux aarch64. On Windows, Rgguf maps
the tensor bytes through a small `CreateFileMapping` shim in place of `mmap`,
while Rggml remains the official parser authority. Rpgen's vendored pgenlib
builds with the same `Makevars.win` pattern that carries pgenlibr on CRAN.

## Correctness discipline

Every quantized codec decoder is pinned **bit-identical** to GGML's reference
(`Rggml_dequantize`, GGML's own type-traits `to_float`) by regression fixtures
in Rgguf and live cross-validation in Rllm - the discipline that caught a real
upstream Q4_K dequantization bug in gguf-tools (~33% wrong decodes of the most
common real-model format; fixed in our vendored copy). The inference graph is
pinned to a pure-R reference forward (~1e-6 on real weights); incremental
KV-cache logits equal whole-batch logits at every position; lossless codecs
round-trip exactly. The integration CI job installs the whole stack and runs
every package's test suite against it (`tests/integration.R`).

## Layout & provenance

```
packages/    the six R packages (independent installable units)
experiments/ falsifiable probes of storage and compute bets
tests/       cross-package integration entry points (run in CI)
tools/       fixture generators and shared scripts
```

The vendored GGML is currently **v0.9.5** (via the CRAN-facing
[`ggmlR`](https://github.com/Zabis13/ggmlR), which supplies the stdio/abort
compliance shim; upstream standalone GGML is at v0.9.11). It is refreshed
deliberately, not continuously: the next refresh is folded into the CUDA work
and tested on real hardware, since the CUDA sources must version-match the
core. See `packages/Rggml/inst/COPYRIGHTS` for the full version & update
policy.

Related but deliberately separate:
[RsimdDispatch](https://github.com/sounkou-bioinfo/RsimdDispatch) (the
runtime-SIMD staging pattern Rggml adopts).

## License

The stack is GPL (>= 2). Rpgen is GPL-3 because it combines PLINK2's GPL-3
import closure with LGPL (>= 3) pgenlib, and RfmallocStatgen is GPL-3 because
its pinned PCAone workhorse is GPL-3. Vendored third-party components (fmalloc,
GGML, pgenlib, PCAone, and ALP) keep their upstream licenses; see each
package's `inst/COPYRIGHTS`.
