---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Rfmalloc

Rfmalloc is an experimental R package for allocating vector storage from a
memory-mapped backing file using a patched copy of
[fmalloc](https://github.com/yasukata/fmalloc). The goal is to explore large,
file-backed R vector storage without routing large allocations back through the
process heap.

The current package exposes ALTREP vectors whose fixed-width atomic payloads and
character-vector string storage are allocated directly from fmalloc. This uses
the ALTREP C entry points available in R's installed headers instead of relying
on the non-API `Rf_allocVector3()` vector allocator hook.

## Current Status

Recent allocator work fixed the stress-path that previously fell back into
ptmalloc/dlmalloc's system `mmap()` path for large requests. Large allocations
are now carved from contiguous runs in the fmalloc backing file instead.

Implemented now:

- ALTREP file-backed allocation for logical, integer, numeric, raw, complex,
  character, and list vectors;
- large allocations spanning multiple fmalloc chunks;
- `malloc`, `free`, and `realloc` support through the patched fmalloc layer;
- reopening backing files created by the current allocator format;
- explicit runtime handles, so multiple fmalloc runtimes can be open in one R
  process;
- explicit runtime modes: `persistent` keeps committed vector payloads in the
  file, while `scratch` returns payloads to fmalloc when ALTREP handles are
  garbage-collected;
- persistent ALTREP serialization for fixed-width atomic and character vectors
  using physical allocation metadata: path, file UUID, type, length, payload
  offset, and byte size;
- ALTREP `Coerce` for supported fmalloc-backed target types, so coercions such
  as `as.numeric(x)` and `as.character(x)` allocate their result in the same
  fmalloc runtime;
- ALTREP `Extract_subset` for vector indexing operations such as `x[i]`,
  returning fmalloc-backed copy results;
- native lifetime tracking from ALTREP vector handles to runtime mappings, so a
  runtime mapping is not destroyed while vectors allocated from it are still
  reachable.

Still experimental / future work:

- view-based ALTREP subsets for simple contiguous or strided indexing patterns;
- a real allocation catalog for listing and validating persistent allocations;
- richer persistence semantics for pointer-containing R types; list vectors are
  session-local containers, although their fmalloc-vector elements can serialize
  through their own persistent references;
- stronger mapped-file lifecycle behavior: clearer close semantics, recovery from
  interrupted sessions, stale backing-file diagnostics, and platform-specific
  mapping edge cases.

## Installation

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("sounkou-bioinfo/Rfmalloc")
```

## Basic Usage

The examples use `local()` and `on.exit()` only to keep temporary backing files
scoped while the README is rendered.

The shortest form uses `init_fmalloc()`, which opens a backing file and stores
it as the package default runtime. Calls to `create_fmalloc_vector()` can then
omit the `runtime` argument:

```{r}
library(Rfmalloc)

local({
  alloc_file <- tempfile(fileext = ".bin")
  init_fmalloc(alloc_file)
  on.exit({
    cleanup_fmalloc()
    unlink(alloc_file)
  }, add = TRUE)

  v_int <- create_fmalloc_vector("integer", 10)
  v_num <- create_fmalloc_vector("numeric", 10)
  v_raw <- create_fmalloc_vector("raw", 4)
  v_cplx <- create_fmalloc_vector("complex", 4)
  v_chr <- create_fmalloc_vector("character", 3)
  v_lst <- create_fmalloc_vector("list", 2)

  v_int[1:3] <- c(1L, 2L, 3L)
  v_num[1:3] <- c(1.1, 2.2, 3.3)
  v_raw[] <- as.raw(1:4)
  v_cplx[] <- c(1+1i, 2+2i, 3+3i, 4+4i)
  v_chr[] <- c("alpha", "beta", NA_character_)
  v_lst[[1]] <- data.frame(x = 1:2)

  list(
    integer = v_int[1:3],
    numeric = v_num[1:3],
    raw = v_raw[],
    complex = v_cplx[],
    character = v_chr[],
    list_first = v_lst[[1]]
  )
})
```

For multiple backing files or explicit lifetime management, prefer runtime
handles returned by `open_fmalloc()`:

```{r}
library(Rfmalloc)

local({
  handle_file <- tempfile(fileext = ".bin")
  rt <- open_fmalloc(handle_file)
  on.exit({
    cleanup_fmalloc(rt)
    unlink(handle_file)
  }, add = TRUE)

  v <- create_fmalloc_vector("integer", 10, runtime = rt)
  v[1:3] <- 10:12
  v[1:3]
})
```

## Larger Allocation Example

This is the kind of stress case the package is meant for: it creates a 5 GB
backing file and allocates about 4 GB of integer payload.

```{r}
library(Rfmalloc)

local({
  large_file <- tempfile(fileext = ".bin")
  init_fmalloc(large_file, size_gb = 5)
  on.exit({
    cleanup_fmalloc()
    unlink(large_file)
  }, add = TRUE)

  # 1 billion integers = about 4 GB of payload, backed by fmalloc.
  # Keep creation and initialization in one local expression so README rendering
  # does not keep an extra transient reference that would force a second 4 GB COW
  # duplicate during `[<-`.
  big_int <- local({
    x <- create_fmalloc_vector("integer", 1e9)
    x[1:5] <- 1:5
    x
  })
  big_int[1:5]
})
```

## Runtime Modes

`open_fmalloc()` and `init_fmalloc()` accept `mode = "persistent"` or
`mode = "scratch"`. The default is `"persistent"`.

- In persistent mode, committed vector payloads are not returned to fmalloc by
  vector finalizers. Fixed-width atomic vectors serialize by reference to the
  physical allocation in the backing file.
- In scratch mode, the backing file is a temporary allocation arena. Vector
  finalizers may return payloads to fmalloc, and serialization falls back to
  ordinary R values.

```{r}
library(Rfmalloc)

local({
  persist_file <- tempfile(fileext = ".bin")
  rt <- open_fmalloc(persist_file, mode = "persistent")
  on.exit({
    cleanup_fmalloc(rt)
    unlink(persist_file)
  }, add = TRUE)

  x <- create_fmalloc_vector("integer", 5, runtime = rt)
  x[] <- 1:5
  roundtrip <- unserialize(serialize(x, NULL))
  roundtrip[]
})
```

## Inspecting fmalloc ALTREP Metadata

R's internal inspector dispatches to the ALTREP `Inspect` hook. Rfmalloc's
inspector reports the vector type, length, payload byte count, runtime mode,
runtime state, payload offset, file UUID, and backing-file path.

```{r}
library(Rfmalloc)

local({
  inspect_file <- tempfile(fileext = ".bin")
  rt <- open_fmalloc(inspect_file, mode = "persistent")
  on.exit({
    cleanup_fmalloc(rt)
    unlink(inspect_file)
  }, add = TRUE)

  inspect_vec <- create_fmalloc_vector("integer", 4, runtime = rt)
  inspect_vec[] <- 1:4
  .Internal(inspect(inspect_vec))
})
```

`inspect()` output is an internal R diagnostic, so exact formatting can vary
between R versions. The `fmalloc_altrep ...` line comes from Rfmalloc's ALTREP
`Inspect` method.

## Character Vectors

Rfmalloc character vectors are ALTSTRING vectors. They do not allocate R
`CHARSXP` objects inside fmalloc. R's current sources keep `CHARSXP` allocation
on internal string-cache paths, and `allocVector3()` is explicitly not a public
API. Instead, Rfmalloc stores string bytes, lengths, encodings, and NA flags in
fmalloc allocations. `STRING_ELT()` materializes an ordinary R `CHARSXP` with
`mkCharLenCE()` when R asks for an element.

```{r}
library(Rfmalloc)

local({
  char_file <- tempfile(fileext = ".bin")
  rt <- open_fmalloc(char_file, mode = "persistent")
  on.exit({
    cleanup_fmalloc(rt)
    unlink(char_file)
  }, add = TRUE)

  chars <- create_fmalloc_vector("character", 3, runtime = rt)
  chars[] <- c("one", NA_character_, "three")

  from_int <- create_fmalloc_vector("integer", 3, runtime = rt)
  from_int[] <- 1:3

  list(
    chars = chars[],
    from_integer = as.character(from_int)[]
  )
})
```

## Multiple Runtimes and Lifetime

Runtime handles make it possible to use more than one backing file in the same R
process. Allocator entry is serialized with native mutexes because the current
fmalloc/ptmalloc layer still has process-global target state internally.

```{r}
library(Rfmalloc)

local({
  file_a <- tempfile(fileext = ".bin")
  file_b <- tempfile(fileext = ".bin")

  rt_a <- open_fmalloc(file_a)
  rt_b <- open_fmalloc(file_b, size_gb = 0.1)

  vec_a <- create_fmalloc_vector("integer", 10, runtime = rt_a)
  vec_b <- create_fmalloc_vector("numeric", 10, runtime = rt_b)
  vec_a[1] <- 101L
  vec_b[1] <- 202

  before_cleanup <- data.frame(
    vector = c("vec_a", "vec_b"),
    value = c(vec_a[1], vec_b[1])
  )

  cleanup_fmalloc(rt_a)
  after_cleanup <- vec_a[1]
  cleanup_fmalloc(rt_b)
  unlink(c(file_a, file_b))

  list(
    before_cleanup = before_cleanup,
    vec_a_after_cleanup = after_cleanup
  )
})
```

Calling `cleanup_fmalloc(rt_a)` marks `rt_a` closed. If vectors from `rt_a` are
still reachable, the native mapping is kept alive until those vectors are
garbage-collected. This is implemented with a native live-vector count and an
ALTREP-held external pointer for each fmalloc-backed vector; no user-visible
attribute is added to the vector.

## Serialization and Persistent References

Rfmalloc has an ALTREP serialization path for persistent runtimes. For
persistent fixed-width atomic and character vectors, `serialize()` stores a small
reference record instead of copying the vector payload into the serialization
stream. The record contains:

- format tag: `"RfmallocRef"`;
- reference format version;
- backing-file path;
- backing-file UUID;
- R vector type;
- vector length;
- payload offset inside the mapped file;
- payload byte size.

During `unserialize()`, Rfmalloc reopens the backing file, verifies the UUID,
checks the recorded dimensions and file bounds, and reconstructs an ALTREP vector
around the same fmalloc allocation.

```{r}
library(Rfmalloc)

local({
  ser_file <- tempfile(fileext = ".bin")
  rt <- open_fmalloc(ser_file, mode = "persistent")

  ints <- create_fmalloc_vector("integer", 4, runtime = rt)
  ints[] <- 101:104
  chars <- create_fmalloc_vector("character", 3, runtime = rt)
  chars[] <- c("alpha", NA_character_, "gamma")

  ints_blob <- serialize(ints, NULL)
  chars_blob <- serialize(chars, NULL)

  cleanup_fmalloc(rt)

  ints_recovered <- unserialize(ints_blob)
  chars_recovered <- unserialize(chars_blob)
  output <- list(
    integers = ints_recovered[],
    characters = chars_recovered[]
  )

  unlink(ser_file)
  output
})
```

Scratch runtimes use ordinary R serialization instead. Their serialized values do
not depend on reopening the scratch backing file.

```{r}
local({
  scratch_file <- tempfile(fileext = ".bin")
  scratch_rt <- open_fmalloc(scratch_file, mode = "scratch")

  scratch_vec <- create_fmalloc_vector("integer", 4, runtime = scratch_rt)
  scratch_vec[] <- 1:4
  scratch_copy <- unserialize(serialize(scratch_vec, NULL))

  cleanup_fmalloc(scratch_rt)
  unlink(scratch_file)
  scratch_copy
})
```

List vectors are serialized as ordinary R list values for now. Persistent list
recovery needs a separate element-level policy because list elements are R
objects, not raw fixed-width payload bytes.

This is reference-based recovery, not name-based object discovery. Rfmalloc does
not yet keep a complete in-file object catalog that can list all allocations or
recover dropped vectors by user names. The serialized reference itself is the
locator for now.

## Watching ALTREP-Controlled Duplication with `lobstr`

`lobstr::obj_addr()` is useful for seeing why the vectors are ALTREP from the
beginning. R's copy-on-write path dispatches to the ALTREP `Duplicate` method,
and the duplicate allocates its payload from fmalloc.

```{r}
library(Rfmalloc)
library(lobstr)

local({
  addr_file <- tempfile(fileext = ".bin")
  init_fmalloc(addr_file)
  on.exit({
    cleanup_fmalloc()
    unlink(addr_file)
  }, add = TRUE)

  cow_a <- create_fmalloc_vector("integer", 10)
  cow_a[1:3] <- 1:3
  cow_b <- cow_a

  before <- data.frame(
    object = c("cow_a", "cow_b"),
    address = c(obj_addr(cow_a), obj_addr(cow_b))
  )

  cow_a[1] <- 99L

  after <- data.frame(
    object = c("cow_a", "cow_b"),
    address = c(obj_addr(cow_a), obj_addr(cow_b))
  )

  list(before = before, after = after)
})
```

The two names start as references to the same `SEXP`. After the write to
`cow_a`, `lobstr::obj_addr()` shows that `cow_a` moved to a new ALTREP object
while `cow_b` still points at the original ALTREP object. That automatic
duplicate is handled by Rfmalloc's ALTREP `Duplicate` method, not by R's ordinary
`allocVector()` path.

`lobstr::obj_addr()` shows the R object header address, not necessarily the data
payload pointer. The payload pointer returned by `DATAPTR()` is a real writable
pointer into the fmalloc mapping for fixed-width atomic types. Character vectors
store string bytes in fmalloc and materialize ordinary R `CHARSXP` values on
`STRING_ELT()` access. List vectors use ALTREP list element methods plus an
R-visible sidecar so the GC can see referenced R objects.

## Why ALTREP Is Required for Copy-on-Write Control

Plain vectors allocated through custom allocator hooks do not control R's
ordinary duplication path. R saves enough allocator state to free those objects,
but ordinary vector duplication does not consult a per-object allocator copy
callback, and the `R_allocator_t::res` field is reserved.

Rfmalloc therefore constructs ALTREP vectors from the beginning. The ALTREP C
header declares a `Duplicate` method slot, so copy-on-write can create another
fmalloc-backed vector instead of falling back to R's ordinary heap allocator.
For all current vector constructors, this also means Rfmalloc no longer needs
the non-API `Rf_allocVector3()` path.

The public ALTREP header used by the installed R on this system exposes
`Get_region` hooks for atomic vector reads and `Set_elt` hooks for strings and
lists, but not an atomic `Set_region` hook. Atomic writes therefore go through
R's normal writable `DATAPTR()` path; internal bulk copies use the direct
fmalloc payload pointer.

## References

- fmalloc: <https://github.com/yasukata/fmalloc>
- Simon Urbanek's mmap allocator example: <https://gist.github.com/s-u/6712c97ca74181f5a1a5>
- UFOs: <https://github.com/PRL-PRG/UFOs>
- Travel: <https://github.com/Jiefei-Wang/Travel>

## License

GPL (>= 2)
