---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
has_bench <- requireNamespace("bench", quietly = TRUE)

library(Rfmalloc)

```

# Rfmalloc

 <!-- badges: start -->
  [![R-CMD-check](https://github.com/sounkou-bioinfo/Rfmalloc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rfmalloc/actions/workflows/R-CMD-check.yaml)
  <!-- badges: end -->

Rfmalloc is an experimental R package for file-backed R vectors using a patched
copy of [fmalloc](https://github.com/yasukata/fmalloc). The current R-facing
vector API creates fmalloc-backed ALTREP vectors for atomic, character, and list
storage.

Earlier prototypes explored R's custom allocator / `Rf_allocVector3()` path (see
Simon Urbanek's [proof-of-concept](https://gist.github.com/s-u/6712c97ca74181f5a1a5)). The
package has moved to direct ALTREP vectors for the public vector API and control over
copy-on-write duplication. The patched
fmalloc `malloc`/`free`/`realloc` layer is still built and installed for native
consumers that use the C API or link against the bundled library.

## Current Status

Implemented now:

- fmalloc-backed ALTREP vectors for logical, integer, numeric, raw, complex,
  character, and list values; list elements are restricted to `NULL` or other
  Rfmalloc vectors from the same runtime;
- explicit runtime handles and `persistent` / `scratch` runtime modes;
- persistent serialization and reopening for fixed-width atomic and character
  vectors;
- fmalloc-backed coercion and subset-copy results for supported vector types;
- nested list-container recovery by reference for persistent runtimes;
- explicit `destroy_fmalloc_vector()` for immediate vector invalidation/cleanup
  with parent-reference safety checks;
- an in-file persistent allocation catalog used to validate serialized
  references;
- an installed C header and `R_RegisterCCallable()` API for other packages.

Known limitations:

- Full operator dispatch rewrites are not yet implemented. Some base `Ops`
  expressions (for example `1 + v`) may materialize ordinary R vectors rather
  than stay within the fmalloc ALTREP payload path.

Still experimental: richer catalog and inspection tooling, and mapped-file
recovery diagnostics.

## Installation

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("sounkou-bioinfo/Rfmalloc")
```

## Basic Usage

The examples use temporary backing files. Several chunks wrap code in `local()`
only so `on.exit()` can clean those files up while the README is rendered; this
is not required in normal interactive use. In your own code, keep the runtime
handle you need and call `cleanup_fmalloc()` when finished.

The shortest form uses `init_fmalloc()`, which opens a backing file and stores
it as the package default runtime. Calls to `create_fmalloc_vector()` can then
omit the `runtime` argument:

```{r}
library(Rfmalloc)
local({
  alloc_file <- tempfile(fileext = ".bin")
  init_fmalloc(alloc_file)
  on.exit({
    cleanup_fmalloc()
    unlink(alloc_file)
  }, add = TRUE)

  v_int <- create_fmalloc_vector("integer", 10)
  v_num <- create_fmalloc_vector("numeric", 10)
  v_raw <- create_fmalloc_vector("raw", 4)
  v_cplx <- create_fmalloc_vector("complex", 4)
  v_chr <- create_fmalloc_vector("character", 3)
  v_lst <- create_fmalloc_vector("list", 2)

  v_int[1:3] <- c(1L, 2L, 3L)
  v_num[1:3] <- c(1.1, 2.2, 3.3)
  v_raw[] <- as.raw(1:4)
  v_cplx[] <- c(1+1i, 2+2i, 3+3i, 4+4i)
  v_chr[] <- c("alpha", "beta", NA_character_)

  list_child <- create_fmalloc_vector("integer", 2)
  list_child[] <- 1:2
  v_lst[[1]] <- list_child

  list(
    integer = v_int[1:3],
    numeric = v_num[1:3],
    raw = v_raw[],
    complex = v_cplx[],
    character = v_chr[],
    list_first = v_lst[[1]][]
  )
})
```

For multiple backing files or explicit lifetime management, prefer runtime
handles returned by `open_fmalloc()`:

```{r}

local({
  handle_file <- tempfile(fileext = ".bin")
  rt <- open_fmalloc(handle_file)
  on.exit({
    cleanup_fmalloc(rt)
    unlink(handle_file)
  }, add = TRUE)

  v <- create_fmalloc_vector("integer", 10, runtime = rt)
  v[1:3] <- 10:12
  v[1:3]
})
```

## Larger Allocation Example

```{r, eval = FALSE}

local({
  large_file <- tempfile(fileext = ".bin")
  init_fmalloc(large_file, size_gb = 10)
  on.exit({
    cleanup_fmalloc()
    unlink(large_file)
  }, add = TRUE)

  # Requires substantial on-disk space; each 2e9-integer vector is ~8 GB payload.
  big_int_a <- create_fmalloc_vector("integer", 2e9)
  big_int_b <- create_fmalloc_vector("integer", 2e9)
  big_int_a[1:5] <- 1:5
  big_int_b[1:5] <- 6:10
  list(
    a_head = big_int_a[1:5],
    b_head = big_int_b[1:5]
  )
})
```

## Explicit destruction and parent safety

`destroy_fmalloc_vector()` immediately releases native bookkeeping for one fmalloc
ALTREP vector and returns a logical indicating whether a live vector was destroyed.
The helper enforces parent-reference safety: a vector cannot be destroyed while it
is still stored as a child of any fmalloc list. When needed, drop parent links
first.

```{r}

local({
  destroy_file <- tempfile(fileext = ".bin")
  rt <- open_fmalloc(destroy_file, mode = "persistent")
  on.exit({
    cleanup_fmalloc(rt)
    unlink(destroy_file)
  }, add = TRUE)

  child <- create_fmalloc_vector("integer", 2, runtime = rt)
  child[] <- 1:2

  parent <- create_fmalloc_vector("list", 1, runtime = rt)
  parent[[1]] <- child

  destroy_error <- try(destroy_fmalloc_vector(child), silent = TRUE)

  parent[[1]] <- NULL
  destroy_ok <- destroy_fmalloc_vector(child)

  list(
    destroy_rejected_with_parent = inherits(destroy_error, "try-error"),
    destroy_succeeded_after_unset = destroy_ok
  )
})
```

In persistent mode, destroying with default semantics retains payload bytes so the
recorded on-disk allocation remains recoverable by normal `serialize()` flows.
Use `destroy_fmalloc_vector(x, unsafe = TRUE)` to reclaim persistent payload
memory and mark the catalog entry as non-recoverable.

This is scoped to the targeted vector(s): if one object is unsafe-destroyed,
other objects in the same runtime remain recoverable.

```{r}

local({
  selective_file <- tempfile(fileext = ".bin")
  rt <- open_fmalloc(selective_file, mode = "persistent")
  on.exit({
    cleanup_fmalloc(rt)
    unlink(selective_file)
  }, add = TRUE)

  keep <- create_fmalloc_vector("integer", 3, runtime = rt)
  drop <- create_fmalloc_vector("character", 2, runtime = rt)
  keep[] <- c(101L, 202L, 303L)
  drop[] <- c("alpha", "beta")

  keep_blob <- serialize(keep, NULL)
  drop_blob <- serialize(drop, NULL)

  destroy_fmalloc_vector(drop, unsafe = TRUE)

  keep_recovered <- unserialize(keep_blob)
  drop_recover_error <- try(unserialize(drop_blob), silent = TRUE)

  list(
    keep_recovered_ok = all.equal(keep_recovered[], c(101L, 202L, 303L)) == TRUE,
    drop_recover_fails = inherits(drop_recover_error, "try-error")
  )
})
```

## Runtime Modes

`open_fmalloc()` and `init_fmalloc()` accept `mode = "persistent"` or
`mode = "scratch"`. The default is `"persistent"`.

- In persistent mode, committed vector payloads are not returned to fmalloc by
  vector finalizers. Fixed-width atomic vectors serialize by reference to the
  physical allocation in the backing file.
- In scratch mode, the backing file is a temporary allocation arena. Vector
  finalizers may return payloads to fmalloc, and serialization falls back to
  ordinary R values.

```{r}

local({
  persist_file <- tempfile(fileext = ".bin")
  rt <- open_fmalloc(persist_file, mode = "persistent")
  on.exit({
    cleanup_fmalloc(rt)
    unlink(persist_file)
  }, add = TRUE)

  x <- create_fmalloc_vector("integer", 5, runtime = rt)
  x[] <- 1:5
  roundtrip <- unserialize(serialize(x, NULL))
  roundtrip[]
})
```

## Inspecting fmalloc ALTREP Metadata

R's internal inspector dispatches to the ALTREP `Inspect` hook. Rfmalloc's
inspector reports the vector type, length, payload byte count, runtime mode,
runtime state, payload offset, file UUID, and backing-file path.

```{r}

local({
  inspect_file <- tempfile(fileext = ".bin")
  rt <- open_fmalloc(inspect_file, mode = "persistent")
  on.exit({
    cleanup_fmalloc(rt)
    unlink(inspect_file)
  }, add = TRUE)

  inspect_vec <- create_fmalloc_vector("integer", 4, runtime = rt)
  inspect_vec[] <- 1:4
  .Internal(inspect(inspect_vec))
})
```

`inspect()` output is an internal R diagnostic, so exact formatting can vary
between R versions. The `fmalloc_altrep ...` line comes from Rfmalloc's ALTREP
`Inspect` method.

## Character Vectors

Rfmalloc character vectors are ALTSTRING vectors. String bytes, lengths,
encodings, and NA flags live in fmalloc storage. R `CHARSXP` values are
materialized on `STRING_ELT()` access; Rfmalloc does not allocate `CHARSXP`
objects inside fmalloc.

```{r}

local({
  char_file <- tempfile(fileext = ".bin")
  rt <- open_fmalloc(char_file, mode = "persistent")
  on.exit({
    cleanup_fmalloc(rt)
    unlink(char_file)
  }, add = TRUE)

  chars <- create_fmalloc_vector("character", 3, runtime = rt)
  chars[] <- c("one", NA_character_, "three")

  from_int <- create_fmalloc_vector("integer", 3, runtime = rt)
  from_int[] <- 1:3

  list(
    chars = chars[],
    from_integer = as.character(from_int)[]
  )
})
```

## Matrix and data.frame constructors and converters

Use constructor helpers for shape-aware allocation, and `as_fmalloc_*()` helpers
to install shape metadata on existing fmalloc-backed vectors:

```{r}

local({
  ctor_file <- tempfile(fileext = ".bin")
  rt <- open_fmalloc(ctor_file, mode = "persistent")
  on.exit({
    cleanup_fmalloc(rt)
    unlink(ctor_file)
  }, add = TRUE)

  m <- create_fmalloc_matrix("integer", nrow = 2L, ncol = 3L, runtime = rt)
  m[] <- 1:6

  base_vec <- create_fmalloc_vector("integer", 6L, runtime = rt)
  base_vec[] <- 11:16
  m_view <- as_fmalloc_matrix(base_vec, ncol = 3L, copy = FALSE)

  # Metadata-only reshape, then mutate through the reshaped object
  m_view[1L, 2L] <- 88L

  a <- create_fmalloc_array("numeric", dim = c(2L, 1L, 3L), runtime = rt)
  a[] <- 1:6

  col_a <- create_fmalloc_vector("integer", 3L, runtime = rt)
  col_b <- create_fmalloc_vector("integer", 3L, runtime = rt)
  col_a[] <- c(1L, 2L, 3L)
  col_b[] <- c(4L, 5L, 6L)
  df <- as_fmalloc_data_frame(a = col_a, b = col_b, stringsAsFactors = FALSE)

  list(
    matrix_dims = dim(m),
    metadata_shares_payload = c(base_vec[3L], m_view[1L, 2L]),
    array_dims = dim(a),
    df_columns = names(df)
  )
})
```

## Multiple Runtimes and Lifetime

Runtime handles make it possible to use more than one backing file in one R
process.

```{r}

local({
  file_a <- tempfile(fileext = ".bin")
  file_b <- tempfile(fileext = ".bin")

  rt_a <- open_fmalloc(file_a)
  rt_b <- open_fmalloc(file_b, size_gb = 0.1)

  vec_a <- create_fmalloc_vector("integer", 10, runtime = rt_a)
  vec_b <- create_fmalloc_vector("numeric", 10, runtime = rt_b)
  vec_a[1] <- 101L
  vec_b[1] <- 202

  before_cleanup <- data.frame(
    vector = c("vec_a", "vec_b"),
    value = c(vec_a[1], vec_b[1])
  )

  cleanup_fmalloc(rt_a)
  after_cleanup <- vec_a[1]
  cleanup_fmalloc(rt_b)
  unlink(c(file_a, file_b))

  list(
    before_cleanup = before_cleanup,
    vec_a_after_cleanup = after_cleanup
  )
})
```

Calling `cleanup_fmalloc(rt_a)` marks `rt_a` closed. If vectors from `rt_a` are
still reachable, the native mapping is kept alive until those vectors are
garbage-collected. This is implemented with a native live-vector count and an
ALTREP-held external pointer for each fmalloc-backed vector; no user-visible
attribute is added to the vector.

## Serialization and Persistent References

Rfmalloc has an ALTREP serialization path for persistent runtimes. For
persistent fixed-width atomic and character vectors, `serialize()` stores a small
reference record instead of copying the vector payload into the serialization
stream. The record contains:

- format tag: `"RfmallocRef"`;
- reference format version;
- backing-file path;
- backing-file UUID;
- R vector type;
- vector length;
- payload offset inside the mapped file;
- payload byte size.

During `unserialize()`, Rfmalloc reopens the backing file, verifies the UUID,
checks the recorded dimensions and file bounds, validates the catalog record and
generation, and reconstructs an ALTREP vector around the same fmalloc allocation.

```{r}

local({
  ser_file <- tempfile(fileext = ".bin")
  rt <- open_fmalloc(ser_file, mode = "persistent")

  ints <- create_fmalloc_vector("integer", 4, runtime = rt)
  ints[] <- 101:104
  chars <- create_fmalloc_vector("character", 3, runtime = rt)
  chars[] <- c("alpha", NA_character_, "gamma")

  catalog <- list_fmalloc_allocations(rt)
  ints_blob <- serialize(ints, NULL)
  chars_blob <- serialize(chars, NULL)

  cleanup_fmalloc(rt)

  ints_recovered <- unserialize(ints_blob)
  chars_recovered <- unserialize(chars_blob)
  output <- list(
    catalog = catalog[, c("record_offset", "generation", "type", "length")],
    integers = ints_recovered[],
    characters = chars_recovered[]
  )

  unlink(ser_file)
  output
})
```

## Catalog diagnostics

Use `diagnose_fmalloc_runtime()` to inspect runtime metadata, the live catalog,
and a compact summary that can help decide when a runtime is a reset candidate:

```{r}
local({
  diag_file <- tempfile(fileext = ".bin")
  rt <- open_fmalloc(diag_file, mode = "persistent")

  a <- create_fmalloc_vector("integer", 4, runtime = rt)
  b <- create_fmalloc_vector("character", 2, runtime = rt)
  a[] <- 1:4
  b[] <- c("left", "right")

  pre <- diagnose_fmalloc_runtime(rt)
  destroy_fmalloc_vector(a)
  post <- diagnose_fmalloc_runtime(rt)

  list(
    mode = post$runtime$mode,
    pre_records = pre$summary$record_count,
    pre_committed = pre$summary$committed_records,
    post_records = post$summary$record_count,
    post_tombstones = post$summary$tombstoned_records,
    compaction_implemented = post$summary$compaction_implemented
  )
})
```

Scratch runtimes use ordinary R serialization instead. Their serialized values do
not depend on reopening the scratch backing file.

```{r}
local({
  scratch_file <- tempfile(fileext = ".bin")
  scratch_rt <- open_fmalloc(scratch_file, mode = "scratch")

  scratch_vec <- create_fmalloc_vector("integer", 4, runtime = scratch_rt)
  scratch_vec[] <- 1:4
  scratch_copy <- unserialize(serialize(scratch_vec, NULL))

  cleanup_fmalloc(scratch_rt)
  unlink(scratch_file)
  scratch_copy
})
```

Rfmalloc list vectors are intentionally not general R lists. Element assignment
accepts only `NULL` or Rfmalloc-backed vectors from the same runtime; ordinary R
objects such as base vectors, data frames, or arbitrary lists are rejected. This
keeps lists inside the same file-backed object universe.

For persistent runtimes, list containers now serialize as reference state when all
children are recoverable fmalloc vectors. The serialized state stores per-slot
child descriptors, so nested list containers can reopen recursively from the same
backing file without carrying session-local `SEXP` pointers.

This is reference-based recovery, not name-based object discovery. The catalog
stores physical allocation records, not user variable names. Serialized
references use the catalog record offset and generation for validation; the
catalog can be listed, but it is not yet a high-level object store for recovering
vectors by name.

## List constraints and recovery examples

List assignment rejects non-fmalloc payloads at runtime.

```{r}

local({
  reject_file <- tempfile(fileext = ".bin")
  rt <- open_fmalloc(reject_file, mode = "persistent")
  on.exit({
    cleanup_fmalloc(rt)
    unlink(reject_file)
  }, add = TRUE)

  fm_list <- create_fmalloc_vector("list", 2, runtime = rt)
  child <- create_fmalloc_vector("integer", 2, runtime = rt)
  child[] <- c(1L, 2L)
  fm_list[[1]] <- child

  bad_assignment <- try({
    fm_list[[2]] <- 1:2
  }, silent = TRUE)

  list(
    same_runtime_child = fm_list[[1]][],
    rejected_non_fmalloc_payload = inherits(bad_assignment, "try-error")
  )
})
```

Nested fmalloc lists recover recursively when serialized and unserialized from a
persistent backing file.

```{r}

local({
  recover_file <- tempfile(fileext = ".bin")
  rt <- open_fmalloc(recover_file, mode = "persistent")
  on.exit(unlink(recover_file), add = TRUE)

  child <- create_fmalloc_vector("integer", 2, runtime = rt)
  child[] <- 1:2

  nested <- create_fmalloc_vector("list", 1, runtime = rt)
  nested[[1]] <- child

  labels <- create_fmalloc_vector("character", 2, runtime = rt)
  labels[] <- c("alpha", "beta")

  fm_list <- create_fmalloc_vector("list", 2, runtime = rt)
  fm_list[[1]] <- nested
  fm_list[[2]] <- labels

  blob <- serialize(fm_list, NULL)

  cleanup_fmalloc(rt)

  recovered <- unserialize(blob)
  list(
    recovered_nested = recovered[[1]][[1]][],
    recovered_labels = recovered[[2]]
  )
})
```

Cross-runtime insertion is also rejected.

```{r}

local({
  cross_file_a <- tempfile(fileext = ".bin")
  cross_file_b <- tempfile(fileext = ".bin")

  rt_a <- open_fmalloc(cross_file_a, mode = "persistent")
  rt_b <- open_fmalloc(cross_file_b, mode = "persistent")
  on.exit({
    cleanup_fmalloc(rt_a)
    cleanup_fmalloc(rt_b)
    unlink(c(cross_file_a, cross_file_b))
  }, add = TRUE)

  child_other_runtime <- create_fmalloc_vector("integer", 2, runtime = rt_b)
  child_other_runtime[] <- c(10L, 20L)

  fm_list <- create_fmalloc_vector("list", 1, runtime = rt_a)
  cross_runtime_error <- try({
    fm_list[[1]] <- child_other_runtime
  }, silent = TRUE)

  list(cross_runtime_rejected = inherits(cross_runtime_error, "try-error"))
})
```

## Performance

Rfmalloc vectors are regular R vectors from R's point of view, but different
operations exercise different paths through R and ALTREP:

- vectorized sequential operations such as `sum(x)` can use contiguous payload
  access;
- scalar loops use ALTREP element access and are more call-heavy;
- `x[i]` creates a fmalloc-backed subset copy for supported vector types;
- `bench::mark()` reports R heap allocation in `mem_alloc`; it does not count
  bytes stored in the fmalloc mapped file.

The small benchmark below uses `bench` and a scratch runtime, so benchmark-only
fmalloc allocations are not preserved as persistent records. Timings are
machine- and R-version-specific; use this as a template for local measurements.

```{r performance-bench, message=FALSE, eval=has_bench}

perf_file <- tempfile(fileext = ".bin")
rt <- open_fmalloc(perf_file, mode = "scratch", size_gb = 0.1)

n <- 100000L
set.seed(1)
idx <- sample.int(n, 2000L)

base_int <- integer(n)
base_int[] <- seq_len(n)

fm_int <- create_fmalloc_vector("integer", n, runtime = rt)
fm_int[] <- base_int

write_env <- new.env(parent = emptyenv())
write_env$base <- base_int
write_env$fm <- create_fmalloc_vector("integer", n, runtime = rt)
write_env$fm[] <- base_int

scalar_sum <- function(x, i) {
  total <- 0L
  for (j in i) total <- total + x[[j]]
  total
}

perf_result <- bench::mark(
  base_sequential_sum = sum(base_int),
  fmalloc_sequential_sum = sum(fm_int),
  base_scalar_read = scalar_sum(base_int, idx),
  fmalloc_scalar_read = scalar_sum(fm_int, idx),
  base_subset_copy = base_int[idx],
  fmalloc_subset_copy = fm_int[idx],
  base_indexed_write = {
    write_env$base[idx] <- 0L
    invisible(write_env$base[1L])
  },
  fmalloc_indexed_write = {
    write_env$fm[idx] <- 0L
    invisible(write_env$fm[1L])
  },
  iterations = 20,
  check = FALSE
)[, c("expression", "median", "itr/sec", "mem_alloc", "gc/sec")]

rm(fm_int, write_env)
invisible(gc())
cleanup_fmalloc(rt)
unlink(perf_file)
perf_result
```

## Native C API for Other Packages

Rfmalloc installs `inst/include/Rfmalloc.h` and registers C-callable entry
points with `R_RegisterCCallable()`. Downstream packages can add Rfmalloc to
`LinkingTo` and `Imports`, include the header, and use the inline wrappers.

The current native surface exposes runtime open/cleanup, vector creation,
default-runtime lookup, catalog listing, runtime diagnostics, and an API-version query. Returned
`SEXP` objects follow normal R API ownership rules.

## References

- fmalloc: <https://github.com/yasukata/fmalloc>
- Simon Urbanek's `Rf_allocVector3()` custom mmap allocator PoC:
  <https://gist.github.com/s-u/6712c97ca74181f5a1a5>
- Rfmalloc's prior custom allocator implementation using `Rf_allocVector3()`
  (for historical comparison):
  https://github.com/sounkou-bioinfo/Rfmalloc/commit/0165953

## License

GPL (>= 2)
