---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Rfmalloc

Rfmalloc is an experimental R package for allocating vector storage from a
memory-mapped backing file using a patched copy of
[fmalloc](https://github.com/yasukata/fmalloc). The goal is to explore large,
file-backed R vector storage without routing large allocations back through the
process heap.

The current package exposes ALTREP vectors whose fixed-width atomic payloads are
allocated directly from fmalloc. This uses the ALTREP C entry points available in
R's installed headers instead of relying on the non-API `Rf_allocVector3()`
vector allocator hook.

## Current Status

Recent allocator work fixed the stress-path that previously fell back into
ptmalloc/dlmalloc's system `mmap()` path for large requests. Large allocations
are now carved from contiguous runs in the fmalloc backing file instead.

Implemented now:

- ALTREP file-backed allocation for logical, integer, numeric, raw, complex,
  character, and list vectors;
- large allocations spanning multiple fmalloc chunks;
- `malloc`, `free`, and `realloc` support through the patched fmalloc layer;
- reopening backing files created by the current allocator format;
- explicit runtime handles, so multiple fmalloc runtimes can be open in one R
  process;
- explicit runtime modes: `persistent` keeps committed vector payloads in the
  file, while `scratch` returns payloads to fmalloc when ALTREP handles are
  garbage-collected;
- persistent ALTREP serialization for fixed-width atomic and character vectors
  using physical allocation metadata: path, file UUID, type, length, payload
  offset, and byte size;
- ALTREP `Coerce` for supported fmalloc-backed target types, so coercions such
  as `as.numeric(x)` and `as.character(x)` allocate their result in the same
  fmalloc runtime;
- native lifetime tracking from ALTREP vector handles to runtime mappings, so a
  runtime mapping is not destroyed while vectors allocated from it are still
  reachable.

Still experimental / future work:

- ALTREP `Extract_subset` methods for operations such as `x[i]` when their
  results should remain fmalloc-backed; today those operations may return
  ordinary R heap vectors;
- a real allocation catalog for listing and validating persistent allocations;
- richer persistence semantics for pointer-containing R types; list vectors are
  session-local containers, although their fmalloc-vector elements can serialize
  through their own persistent references;
- stronger mapped-file lifecycle behavior: clearer close semantics, recovery from
  interrupted sessions, stale backing-file diagnostics, and platform-specific
  mapping edge cases.

## Installation

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("sounkou-bioinfo/Rfmalloc")
```

## Basic Usage

The shortest form uses `init_fmalloc()`, which opens a backing file and stores
it as the package default runtime. Calls to `create_fmalloc_vector()` can then
omit the `runtime` argument:

```{r}
library(Rfmalloc)

alloc_file <- tempfile(fileext = ".bin")
init_fmalloc(alloc_file)

v_int <- create_fmalloc_vector("integer", 10)
v_num <- create_fmalloc_vector("numeric", 10)
v_raw <- create_fmalloc_vector("raw", 4)
v_cplx <- create_fmalloc_vector("complex", 4)
v_chr <- create_fmalloc_vector("character", 3)
v_lst <- create_fmalloc_vector("list", 2)

v_int[1:3] <- c(1L, 2L, 3L)
v_num[1:3] <- c(1.1, 2.2, 3.3)
v_raw[] <- as.raw(1:4)
v_cplx[] <- c(1+1i, 2+2i, 3+3i, 4+4i)
v_chr[] <- c("alpha", "beta", NA_character_)
v_lst[[1]] <- data.frame(x = 1:2)

v_int[1:3]
v_num[1:3]
v_raw
v_cplx
v_chr
v_lst

rm(v_int, v_num, v_raw, v_cplx, v_chr, v_lst)
gc()
cleanup_fmalloc()
unlink(alloc_file)
```

For multiple backing files or explicit lifetime management, prefer runtime
handles returned by `open_fmalloc()`:

```{r}
library(Rfmalloc)

handle_file <- tempfile(fileext = ".bin")
rt <- open_fmalloc(handle_file)

v <- create_fmalloc_vector("integer", 10, runtime = rt)
v[1:3] <- 10:12
v[1:3]

rm(v)
gc()
cleanup_fmalloc(rt)
unlink(handle_file)
```

## Larger Allocation Example

This is the kind of stress case the package is meant for: it creates a 5 GB
backing file and allocates about 4 GB of integer payload.

```{r}
library(Rfmalloc)

large_file <- tempfile(fileext = ".bin")
init_fmalloc(large_file, size_gb = 5)

# 1 billion integers = about 4 GB of payload, backed by fmalloc.
# Keep creation and initialization in one local expression so README rendering
# does not keep an extra transient reference that would force a second 4 GB COW
# duplicate during `[<-`.
big_int <- local({
  x <- create_fmalloc_vector("integer", 1e9)
  x[1:5] <- 1:5
  x
})
big_int[1:5]

rm(big_int)
gc()
cleanup_fmalloc()
unlink(large_file)
```

## Runtime Modes

`open_fmalloc()` and `init_fmalloc()` accept `mode = "persistent"` or
`mode = "scratch"`. The default is `"persistent"`.

- In persistent mode, committed vector payloads are not returned to fmalloc by
  vector finalizers. Fixed-width atomic vectors serialize by reference to the
  physical allocation in the backing file.
- In scratch mode, the backing file is a temporary allocation arena. Vector
  finalizers may return payloads to fmalloc, and serialization falls back to
  ordinary R values.

```{r}
library(Rfmalloc)

persist_file <- tempfile(fileext = ".bin")
rt <- open_fmalloc(persist_file, mode = "persistent")

x <- create_fmalloc_vector("integer", 5, runtime = rt)
x[] <- 1:5
roundtrip <- unserialize(serialize(x, NULL))
roundtrip[]

rm(x, roundtrip)
gc()
cleanup_fmalloc(rt)
unlink(persist_file)
```

## Inspecting fmalloc ALTREP Metadata

R's internal inspector dispatches to the ALTREP `Inspect` hook. Rfmalloc's
inspector reports the vector type, length, payload byte count, runtime mode,
runtime state, payload offset, file UUID, and backing-file path.

```{r}
library(Rfmalloc)

inspect_file <- tempfile(fileext = ".bin")
rt <- open_fmalloc(inspect_file, mode = "persistent")

inspect_vec <- create_fmalloc_vector("integer", 4, runtime = rt)
inspect_vec[] <- 1:4
.Internal(inspect(inspect_vec))

rm(inspect_vec)
gc()
cleanup_fmalloc(rt)
unlink(inspect_file)
```

`inspect()` output is an internal R diagnostic, so exact formatting can vary
between R versions. The `fmalloc_altrep ...` line comes from Rfmalloc's ALTREP
`Inspect` method.

## Character Vectors

Rfmalloc character vectors are ALTSTRING vectors. They do not allocate R
`CHARSXP` objects inside fmalloc. R's current sources keep `CHARSXP` allocation
on internal string-cache paths, and `allocVector3()` is explicitly not a public
API. Instead, Rfmalloc stores string bytes, lengths, encodings, and NA flags in
fmalloc allocations. `STRING_ELT()` materializes an ordinary R `CHARSXP` with
`mkCharLenCE()` when R asks for an element.

```{r}
library(Rfmalloc)

char_file <- tempfile(fileext = ".bin")
rt <- open_fmalloc(char_file, mode = "persistent")

chars <- create_fmalloc_vector("character", 3, runtime = rt)
chars[] <- c("one", NA_character_, "three")
chars[]

from_int <- create_fmalloc_vector("integer", 3, runtime = rt)
from_int[] <- 1:3
as.character(from_int)[]

rm(chars, from_int)
gc()
cleanup_fmalloc(rt)
unlink(char_file)
```

## Multiple Runtimes and Lifetime

Runtime handles make it possible to use more than one backing file in the same R
process. Allocator entry is serialized with native mutexes because the current
fmalloc/ptmalloc layer still has process-global target state internally.

```{r}
library(Rfmalloc)

file_a <- tempfile(fileext = ".bin")
file_b <- tempfile(fileext = ".bin")

rt_a <- open_fmalloc(file_a)
rt_b <- open_fmalloc(file_b, size_gb = 0.1)

vec_a <- create_fmalloc_vector("integer", 10, runtime = rt_a)
vec_b <- create_fmalloc_vector("numeric", 10, runtime = rt_b)
vec_a[1] <- 101L
vec_b[1] <- 202

data.frame(
  vector = c("vec_a", "vec_b"),
  value = c(vec_a[1], vec_b[1])
)

cleanup_fmalloc(rt_a)
vec_a[1]

rm(vec_a)
gc()
cleanup_fmalloc(rt_b)
rm(vec_b)
gc()
unlink(c(file_a, file_b))
```

Calling `cleanup_fmalloc(rt_a)` marks `rt_a` closed. If vectors from `rt_a` are
still reachable, the native mapping is kept alive until those vectors are
garbage-collected. This is implemented with a native live-vector count and an
ALTREP-held external pointer for each fmalloc-backed vector; no user-visible
attribute is added to the vector.

## Reopening and Persistent References

A backing file can be reopened. Persistent-mode fixed-width atomic and character
ALTREP vectors can also be serialized by reference: the serialized state stores
the backing path, file UUID, type, length, payload offset, and byte size. During
unserialization, Rfmalloc reopens the file, verifies the UUID, and reconstructs
an ALTREP around the same allocation.

```{r}
library(Rfmalloc)

reopen_file <- tempfile(fileext = ".bin")

rt <- open_fmalloc(reopen_file, mode = "persistent")
first_vec <- create_fmalloc_vector("integer", 100, runtime = rt)
first_vec[1:3] <- 1:3
blob <- serialize(first_vec, NULL)

rm(first_vec)
gc()
cleanup_fmalloc(rt)
rm(rt)
gc()

recovered <- unserialize(blob)
recovered[1:3]
typeof(recovered)

rm(recovered)
gc()
unlink(reopen_file)
```

This is reference-based recovery, not name-based object discovery. Rfmalloc does
not yet keep a complete in-file object catalog that can list all allocations or
recover dropped vectors by user names. The serialized reference itself is the
locator for now.

## Watching ALTREP-Controlled Duplication with `lobstr`

`lobstr::obj_addr()` is useful for seeing why the vectors are ALTREP from the
beginning. R's copy-on-write path dispatches to the ALTREP `Duplicate` method,
and the duplicate allocates its payload from fmalloc.

```{r}
library(Rfmalloc)
library(lobstr)

addr_file <- tempfile(fileext = ".bin")
init_fmalloc(addr_file)

cow_a <- create_fmalloc_vector("integer", 10)
cow_a[1:3] <- 1:3
cow_b <- cow_a

before <- data.frame(
  object = c("cow_a", "cow_b"),
  address = c(obj_addr(cow_a), obj_addr(cow_b))
)

cow_a[1] <- 99L

after <- data.frame(
  object = c("cow_a", "cow_b"),
  address = c(obj_addr(cow_a), obj_addr(cow_b))
)

before
after

rm(cow_a, cow_b)
gc()
cleanup_fmalloc()
unlink(addr_file)
```

The two names start as references to the same `SEXP`. After the write to
`cow_a`, `lobstr::obj_addr()` shows that `cow_a` moved to a new ALTREP object
while `cow_b` still points at the original ALTREP object. That automatic
duplicate is handled by Rfmalloc's ALTREP `Duplicate` method, not by R's ordinary
`allocVector()` path.

`lobstr::obj_addr()` shows the R object header address, not necessarily the data
payload pointer. The payload pointer returned by `DATAPTR()` is a real writable
pointer into the fmalloc mapping for fixed-width atomic types. Character vectors
store string bytes in fmalloc and materialize ordinary R `CHARSXP` values on
`STRING_ELT()` access. List vectors use ALTREP list element methods plus an
R-visible sidecar so the GC can see referenced R objects.

## Why ALTREP Is Required for Copy-on-Write Control

Plain vectors allocated through custom allocator hooks do not control R's
ordinary duplication path. R saves enough allocator state to free those objects,
but ordinary vector duplication does not consult a per-object allocator copy
callback, and the `R_allocator_t::res` field is reserved.

Rfmalloc therefore constructs ALTREP vectors from the beginning. The ALTREP C
header declares a `Duplicate` method slot, so copy-on-write can create another
fmalloc-backed vector instead of falling back to R's ordinary heap allocator.
For all current vector constructors, this also means Rfmalloc no longer needs
the non-API `Rf_allocVector3()` path.

The public ALTREP header used by the installed R on this system exposes
`Get_region` hooks for atomic vector reads and `Set_elt` hooks for strings and
lists, but not an atomic `Set_region` hook. Atomic writes therefore go through
R's normal writable `DATAPTR()` path; internal bulk copies use the direct
fmalloc payload pointer.

## References

- fmalloc: <https://github.com/yasukata/fmalloc>
- Simon Urbanek's mmap allocator example: <https://gist.github.com/s-u/6712c97ca74181f5a1a5>
- UFOs: <https://github.com/PRL-PRG/UFOs>
- Travel: <https://github.com/Jiefei-Wang/Travel>

## License

GPL (>= 2)
