---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Rfmalloc

Rfmalloc is an experimental R package for allocating vector storage from a
memory-mapped backing file using a patched copy of
[fmalloc](https://github.com/yasukata/fmalloc). The goal is to explore large,
file-backed R vector storage without routing large allocations back through the
process heap.

The current package exposes an `Rf_allocVector3()`-based custom allocator for
integer, numeric, and logical vectors. This is useful for experiments and stress
testing, but `Rf_allocVector3()` is not part of R's public API, so this package
should still be treated as research software.

## Current Status

Recent allocator work fixed the stress-path that previously fell back into
ptmalloc/dlmalloc's system `mmap()` path for large requests. Large allocations
are now carved from contiguous runs in the fmalloc backing file instead.

Implemented now:

- file-backed allocation for integer, numeric, and logical vectors;
- large allocations spanning multiple fmalloc chunks;
- `malloc`, `free`, and `realloc` support through the patched fmalloc layer;
- reopening backing files created by the current allocator format.

Still experimental / future work:

- ALTREP wrappers so R-level duplication, subsetting, and copy-on-write can stay
  file-backed;
- richer persistence semantics and metadata;
- support for more vector types;
- production-grade lifecycle management for mapped files.

## Installation

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("sounkou-bioinfo/Rfmalloc")
```

## Basic Usage

```{r}
library(Rfmalloc)

alloc_file <- tempfile(fileext = ".bin")
init_fmalloc(alloc_file)

v_int <- create_fmalloc_vector("integer", 10)
v_num <- create_fmalloc_vector("numeric", 10)

v_int[1:3] <- c(1L, 2L, 3L)
v_num[1:3] <- c(1.1, 2.2, 3.3)

v_int[1:3]
v_num[1:3]

rm(v_int, v_num)
gc()
cleanup_fmalloc()
unlink(alloc_file)
```

## Larger Allocation Example

```{r}
library(Rfmalloc)

large_file <- tempfile(fileext = ".bin")
init_fmalloc(large_file, size_gb = 0.1)

# About 20 MB of integer payload, larger than the historical 16 MB chunk limit.
big_int <- create_fmalloc_vector("integer", 5e6)
big_int[1:5] <- 1:5
big_int[1:5]

rm(big_int)
gc()
cleanup_fmalloc()
unlink(large_file)
```

## Reopening a Backing File

The allocator metadata lives in the mapped file, so a backing file created by
the current format can be closed and reopened. Today this demonstrates allocator
state reuse; recovering named R objects still needs a higher-level root-object or
ALTREP layer.

```{r}
library(Rfmalloc)

reopen_file <- tempfile(fileext = ".bin")

first_init <- init_fmalloc(reopen_file)
first_vec <- create_fmalloc_vector("integer", 100)
first_vec[1:3] <- 1:3
first_init

rm(first_vec)
gc()
cleanup_fmalloc()

second_init <- init_fmalloc(reopen_file)
second_vec <- create_fmalloc_vector("numeric", 100)
second_vec[1:3] <- c(10.5, 20.5, 30.5)
second_init
second_vec[1:3]

rm(second_vec)
gc()
cleanup_fmalloc()
unlink(reopen_file)
```

## Watching R's Duplication with `lobstr`

`lobstr::obj_addr()` and R's internal inspector are useful for seeing why the
current custom-allocator approach is not enough. The initial vector is backed by
fmalloc, but when R performs copy-on-write it duplicates through R's ordinary
vector allocation path.

```{r}
library(Rfmalloc)
library(lobstr)

addr_file <- tempfile(fileext = ".bin")
init_fmalloc(addr_file)

cow_a <- create_fmalloc_vector("integer", 10)
cow_a[1:3] <- 1:3
cow_b <- cow_a

data.frame(
  object = c("cow_a", "cow_b"),
  address = c(obj_addr(cow_a), obj_addr(cow_b))
)
capture.output(.Internal(inspect(cow_a)))[1]

tracemem(cow_a)
cow_a[1] <- 99L

data.frame(
  object = c("cow_a", "cow_b"),
  address = c(obj_addr(cow_a), obj_addr(cow_b))
)
capture.output(.Internal(inspect(cow_a)))[1]
capture.output(.Internal(inspect(cow_b)))[1]

rm(cow_a, cow_b)
gc()
cleanup_fmalloc()
unlink(addr_file)
```

The two names start as references to the same `SEXP`. After the write to
`cow_a`, `tracemem()` reports a duplicate and `lobstr::obj_addr()` shows that
`cow_a` moved to a new object while `cow_b` still points at the original object.
That automatic duplicate is made by R's normal `allocVector()` path, not by the
fmalloc allocator used for the original `Rf_allocVector3()` call.

`lobstr::obj_addr()` shows the R object header address, not necessarily the data
payload pointer. For allocator debugging we may add a small `.Call()` helper that
prints both the `SEXP` address and `DATAPTR()` and whether the payload lies
inside the current fmalloc mapping.

## Why ALTREP Is Still the Likely Next Step

Allocating the initial vector storage in the backing file is not enough to make
all R operations file-backed. R's normal duplication paths often call
`allocVector()`, not `Rf_allocVector3()`, so copy-on-write or subsetting may
produce ordinary heap-backed vectors.

We cannot already pass an automatic duplication method through the current
standard-vector custom allocator. R saves the custom allocator so the original
object can be freed correctly, but ordinary vector duplication does not consult a
per-object allocator hook. The `R_allocator_t::res` field is reserved and is not
a supported copy callback.

A UFO/Travel-like direction is still interesting, but for this package the next
practical layer is likely ALTREP: implement `Dataptr`, `Elt`, `Get_region`, and
especially `Duplicate` so R can preserve file-backed storage during ordinary
vector operations. The ALTREP `Duplicate` method is the supported place where we
can allocate the copied vector through fmalloc as well. Pointer-access
instrumentation is only needed if we want more transparent interception than
ALTREP provides.

## References

- fmalloc: <https://github.com/yasukata/fmalloc>
- Simon Urbanek's mmap allocator example: <https://gist.github.com/s-u/6712c97ca74181f5a1a5>
- UFOs: <https://github.com/PRL-PRG/UFOs>
- Travel: <https://github.com/Jiefei-Wang/Travel>

## License

GPL (>= 2)
