---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Rfmalloc — out-of-core arrays for R

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/Rfmalloc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rfmalloc/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

**One repo, one story: array computation on data that does not fit in RAM,
over pluggable storage codecs and compute backends.**

Memory-mapped file storage (a patched [fmalloc](https://github.com/yasukata/fmalloc)
allocator behind R's ALTREP) turns RAM from a hard cutoff into a spectrum of
speed levels. On top of that substrate sit two plugin registries that everything
else closes over:

- a **codec registry** — how bytes are stored: lossless (`f64` `f32` `f16`
  `bf16` `alp` `sparse`) for scientific data where bit-exactness is
  non-negotiable, and lossy GGUF quantized blocks (`q4_0` `q4_1` `q8_0`
  `q2_k` `q4_k` `q6_k`) for model weights at 2–8 bits each;
- a **matmul backend registry** — how products are computed: R's BLAS by
  default, [GGML](https://github.com/ggml-org/ggml) natively in quantized
  space, GPU backends next.

The contract that makes it compose is the **decline protocol**: any backend may
refuse any product, and the substrate falls back to bounded decode panels +
BLAS. Plugins change how fast you get an answer, never whether it is correct.

## The packages

| package | role |
|---|---|
| [`packages/Rfmalloc`](packages/Rfmalloc) | the substrate: file-backed ALTREP vectors/matrices/tensors, codec + backend registries, panel matmul, out-of-core eviction, PCA/rowVars for genomics-scale matrices |
| [`packages/Rgguf`](packages/Rgguf) | GGUF model files: metadata, tensor directory, dequantized or **native (still-quantized)** imports into fmalloc storage |
| [`packages/Rggml`](packages/Rggml) | compute carrier: vendored GGML with a BLAS backend (R's own BLAS via a `cblas`→Fortran shim) and runtime-SIMD-dispatched quantized kernels (AVX2/NEON, no non-portable-flags NOTE) |
| [`packages/Rllm`](packages/Rllm) | the composition layer: registers Rggml as Rfmalloc's codec-aware matmul backend, so `dense %*% quantized_tensor` runs natively in quantized space, zero-copy from the mmap'd payload |

The whole stack, live — weights quantized to 4.5 bits/weight into mmap'd
storage, multiplied by GGML's SIMD kernels without ever decoding to double
(this chunk runs when the README is rendered):

```{r stack}
library(Rllm)   # registers + selects the ggml backend on load

rt <- Rfmalloc::open_fmalloc(tempfile(fileext = ".bin"))
set.seed(1)
W  <- matrix(rnorm(512 * 8, sd = 0.4), nrow = 512)   # a "weight" matrix
Wq <- rllm_quantize_tensor(W, "q4_k", runtime = rt)  # 4.5 bits/weight, file-backed
Wq

X <- matrix(rnorm(4 * 512), nrow = 4)                # "activations"
Y <- X %*% Wq                                        # native quantized product
# faithful to the full-precision product, at ~1/14th the weight bytes:
sqrt(sum((Y - X %*% W)^2) / sum((X %*% W)^2))
```

On a real model the weights come straight off the GGUF file instead:

```{r gguf, eval = FALSE}
W <- Rgguf::gguf_tensor("model.gguf", "blk.0.ffn_up.weight",
                        runtime = rt, as = "native")  # zero-copy, stays q4_k
Y <- X %*% W
```

## Install

From the [r-universe](https://sounkou-bioinfo.r-universe.dev):

```{r install, eval = FALSE}
install.packages("Rllm",
  repos = c("https://sounkou-bioinfo.r-universe.dev", getOption("repos")))
```

or from GitHub via subdir refs:

```{r pak, eval = FALSE}
pak::pak("sounkou-bioinfo/Rfmalloc/packages/Rllm")
```

Unix (Linux/macOS) only, except `Rfmalloc` itself which also builds on Windows.

## Correctness discipline

Every quantized codec decoder is pinned **bit-identical** to GGML's reference
(`Rggml_dequantize`, GGML's own type-traits `to_float`) by regression fixtures
in Rgguf and live cross-validation in Rllm — a discipline that caught a real
upstream Q4_K dequantization bug in gguf-tools (~33% wrong decodes of the most
common real-model format; fixed in our vendored copy). Lossless codecs
round-trip exactly by construction. The integration CI job installs the whole
stack and runs every package's test suite against it (`tests/integration.R`).

## Layout

```
packages/    the four R packages (independent installable units)
tests/       cross-package integration entry points (run in CI)
tools/       fixture generators and shared scripts
```

Related but deliberately separate:
[RsimdDispatch](https://github.com/sounkou-bioinfo/RsimdDispatch) (the
runtime-SIMD staging pattern Rggml adopts, useful to any R package).

## License

GPL (>= 2). Vendored third-party components (fmalloc, gguflib, GGML via ggmlR)
keep their upstream licenses; see each package's `inst/COPYRIGHTS`.
