---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Rfmalloc

Rfmalloc is an experimental R package for allocating vector storage from a
memory-mapped backing file using a patched copy of
[fmalloc](https://github.com/yasukata/fmalloc). The goal is to explore large,
file-backed R vector storage without routing large allocations back through the
process heap.

The current package exposes ALTREP vectors whose fixed-width atomic payloads are
allocated directly from fmalloc. This keeps the implementation on R's documented
C/ALTREP surface instead of relying on the non-API `Rf_allocVector3()` vector
allocator hook.

## Current Status

Recent allocator work fixed the stress-path that previously fell back into
ptmalloc/dlmalloc's system `mmap()` path for large requests. Large allocations
are now carved from contiguous runs in the fmalloc backing file instead.

Implemented now:

- ALTREP file-backed allocation for logical, integer, numeric, raw, complex,
  character, and list vectors;
- large allocations spanning multiple fmalloc chunks;
- `malloc`, `free`, and `realloc` support through the patched fmalloc layer;
- reopening backing files created by the current allocator format;
- explicit runtime handles, so multiple fmalloc runtimes can be open in one R
  process;
- native lifetime tracking from ALTREP vector handles to runtime mappings, so a
  runtime mapping is not destroyed while vectors allocated from it are still
  reachable.

Still experimental / future work:

- ALTREP `Extract_subset` and `Coerce` methods for operations such as `x[i]`
  and `as.numeric(x)` when their results should remain fmalloc-backed; today
  those operations may return ordinary R heap vectors;
- richer persistence semantics and metadata;
- richer persistence semantics for pointer-containing R types; character and
  list vectors are session-local because their elements are R object pointers;
- production-grade lifecycle management for mapped files.

## Installation

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("sounkou-bioinfo/Rfmalloc")
```

## Basic Usage

The shortest form uses `init_fmalloc()`, which opens a backing file and stores
it as the package default runtime. Calls to `create_fmalloc_vector()` can then
omit the `runtime` argument:

```{r}
library(Rfmalloc)

alloc_file <- tempfile(fileext = ".bin")
init_fmalloc(alloc_file)

v_int <- create_fmalloc_vector("integer", 10)
v_num <- create_fmalloc_vector("numeric", 10)
v_raw <- create_fmalloc_vector("raw", 4)
v_cplx <- create_fmalloc_vector("complex", 4)
v_chr <- create_fmalloc_vector("character", 3)
v_lst <- create_fmalloc_vector("list", 2)

v_int[1:3] <- c(1L, 2L, 3L)
v_num[1:3] <- c(1.1, 2.2, 3.3)
v_raw[] <- as.raw(1:4)
v_cplx[] <- c(1+1i, 2+2i, 3+3i, 4+4i)
v_chr[] <- c("a", "b", "c")
v_lst[[1]] <- data.frame(x = 1:2)

v_int[1:3]
v_num[1:3]
v_raw
v_cplx
v_chr
v_lst

rm(v_int, v_num, v_raw, v_cplx, v_chr, v_lst)
gc()
cleanup_fmalloc()
unlink(alloc_file)
```

For multiple backing files or explicit lifetime management, prefer runtime
handles returned by `open_fmalloc()`:

```{r}
library(Rfmalloc)

handle_file <- tempfile(fileext = ".bin")
rt <- open_fmalloc(handle_file)

v <- create_fmalloc_vector("integer", 10, runtime = rt)
v[1:3] <- 10:12
v[1:3]

rm(v)
gc()
cleanup_fmalloc(rt)
unlink(handle_file)
```

## Larger Allocation Example

```{r}
library(Rfmalloc)

large_file <- tempfile(fileext = ".bin")
init_fmalloc(large_file, size_gb = 0.1)

# About 20 MB of integer payload, larger than the historical 16 MB chunk limit.
big_int <- create_fmalloc_vector("integer", 5e6)
big_int[1:5] <- 1:5
big_int[1:5]

rm(big_int)
gc()
cleanup_fmalloc()
unlink(large_file)
```

## Multiple Runtimes and Lifetime

Runtime handles make it possible to use more than one backing file in the same R
process. Allocator entry is serialized with native mutexes because the current
fmalloc/ptmalloc layer still has process-global target state internally.

```{r}
library(Rfmalloc)

file_a <- tempfile(fileext = ".bin")
file_b <- tempfile(fileext = ".bin")

rt_a <- open_fmalloc(file_a)
rt_b <- open_fmalloc(file_b, size_gb = 0.1)

vec_a <- create_fmalloc_vector("integer", 10, runtime = rt_a)
vec_b <- create_fmalloc_vector("numeric", 10, runtime = rt_b)
vec_a[1] <- 101L
vec_b[1] <- 202

data.frame(
  vector = c("vec_a", "vec_b"),
  value = c(vec_a[1], vec_b[1])
)

cleanup_fmalloc(rt_a)
vec_a[1]

rm(vec_a)
gc()
cleanup_fmalloc(rt_b)
rm(vec_b)
gc()
unlink(c(file_a, file_b))
```

Calling `cleanup_fmalloc(rt_a)` marks `rt_a` closed. If vectors from `rt_a` are
still reachable, the native mapping is kept alive until those vectors are
garbage-collected. This is implemented with a native live-vector count and an
ALTREP-held external pointer for each fmalloc-backed vector; no user-visible
attribute is added to the vector.

## Reopening Is Not Vector Recovery Yet

A backing file can be reopened, but this currently reopens only the allocator
state. We are **not** getting the R vector object back yet. The package does not
currently store a root table such as `(name, type, length, offset)`, so an
ALTREP vector cannot yet be safely reconstructed after it has been dropped by
R's garbage collector.

```{r}
library(Rfmalloc)

reopen_file <- tempfile(fileext = ".bin")

first_init <- init_fmalloc(reopen_file)
first_vec <- create_fmalloc_vector("integer", 100)
first_vec[1:3] <- 1:3
first_init
first_vec[1:3]

rm(first_vec)
gc()
cleanup_fmalloc()

second_init <- init_fmalloc(reopen_file)
second_init
exists("first_vec")

cleanup_fmalloc()
unlink(reopen_file)
```

The `FALSE` from `second_init` tells us the file already contained current
fmalloc allocator metadata. It does **not** mean `first_vec` was recovered. Real
object recovery needs a higher-level object catalog and likely an ALTREP vector
constructor that can recreate an R vector from `(file, offset, type, length)`.

## Watching ALTREP-Controlled Duplication with `lobstr`

`lobstr::obj_addr()` and R's internal inspector are useful for seeing why the
vectors are ALTREP from the beginning. R's copy-on-write path dispatches to the
ALTREP `Duplicate` method, and the duplicate allocates its payload from fmalloc.

```{r}
library(Rfmalloc)
library(lobstr)

addr_file <- tempfile(fileext = ".bin")
init_fmalloc(addr_file)

cow_a <- create_fmalloc_vector("integer", 10)
cow_a[1:3] <- 1:3
cow_b <- cow_a

data.frame(
  object = c("cow_a", "cow_b"),
  address = c(obj_addr(cow_a), obj_addr(cow_b))
)
capture.output(.Internal(inspect(cow_a)))[1]

tracemem(cow_a)
cow_a[1] <- 99L

data.frame(
  object = c("cow_a", "cow_b"),
  address = c(obj_addr(cow_a), obj_addr(cow_b))
)
capture.output(.Internal(inspect(cow_a)))[1]
capture.output(.Internal(inspect(cow_b)))[1]

rm(cow_a, cow_b)
gc()
cleanup_fmalloc()
unlink(addr_file)
```

The two names start as references to the same `SEXP`. After the write to
`cow_a`, `tracemem()` reports a duplicate and `lobstr::obj_addr()` shows that
`cow_a` moved to a new ALTREP object while `cow_b` still points at the original
ALTREP object. That automatic duplicate is handled by Rfmalloc's ALTREP
`Duplicate` method, not by R's ordinary `allocVector()` path.

`lobstr::obj_addr()` shows the R object header address, not necessarily the data
payload pointer. The payload pointer returned by `DATAPTR()` is a real writable
pointer into the fmalloc mapping for fixed-width atomic types. Character and list
vectors use ALTREP string/list element methods plus an R-visible sidecar so the
GC can see referenced R objects.

## Why ALTREP Is Required for Copy-on-Write Control

Plain vectors allocated through custom allocator hooks do not control R's
ordinary duplication path. R saves enough allocator state to free those objects,
but ordinary vector duplication does not consult a per-object allocator copy
callback, and the `R_allocator_t::res` field is reserved.

Rfmalloc therefore constructs ALTREP vectors from the beginning. ALTREP gives the
package a supported `Duplicate` method, so copy-on-write can create another
fmalloc-backed vector instead of falling back to R's ordinary heap allocator.
For all current vector constructors, this also means Rfmalloc no longer needs
the non-API `Rf_allocVector3()` path.

## References

- fmalloc: <https://github.com/yasukata/fmalloc>
- Simon Urbanek's mmap allocator example: <https://gist.github.com/s-u/6712c97ca74181f5a1a5>
- UFOs: <https://github.com/PRL-PRG/UFOs>
- Travel: <https://github.com/Jiefei-Wang/Travel>

## License

GPL (>= 2)
