---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Rllm

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/Rfmalloc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rfmalloc/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Rllm is the **composition layer** of the
[Rfmalloc](https://github.com/sounkou-bioinfo/Rfmalloc) out-of-core array
stack: it wires [Rfmalloc](../Rfmalloc) (file-backed, memory-mapped storage),
[Rgguf](../Rgguf) (GGUF model files as typed storage views) and
[Rggml](../Rggml) (vendored GGML compute with runtime-SIMD-dispatched
quantized kernels) into LLM
inference - quantized weights stay in their GGUF encoding, memory-mapped from
disk, and are never decoded to double on the compute path.

The first composition is a codec-aware matrix-product backend:
`dense %*% quantized_fmalloc_tensor` contracts the encoded blocks directly
through GGML after registering with Rfmalloc's backend registry. The second is
a llama-architecture inference engine: GGUF loading, a forward graph, KV
caching and byte-level generation over the same typed storage.

## Quantized products, zero-copy

```{r bridge}
library(Rllm)   # registers + selects the ggml backend

rt <- Rfmalloc::open_fmalloc(tempfile(fileext = ".bin"))
set.seed(1)
W  <- matrix(rnorm(512 * 8, sd = 0.4), nrow = 512)
Wq <- rllm_quantize_tensor(W, "q4_k", runtime = rt)  # 4.5 bits/weight, mmap'd
X  <- matrix(rnorm(4 * 512), nrow = 4)
Y  <- X %*% Wq                                       # GGML SIMD, no f64 decode
sqrt(sum((Y - X %*% W)^2) / sum((X %*% W)^2))        # q4_k-accurate
```

## Inference: bytes in, bytes out

The model I/O boundary is **token ids and raw bytes** - never character
vectors in the core. `rllm_encode()`/`rllm_decode()` are edge codecs built
from the GGUF's own byte-level BPE metadata (no external tokenizer);
`rawToChar()` is the caller's interpretation. The same boundary serves
byte-level, vision, or pure-codec models later.

```{r infer, eval = FALSE}
rt <- Rfmalloc::open_fmalloc("work.bin", size_gb = 0.5)    # outputs and KV cache
model <- rllm_gguf_model("SmolLM2-135M.Q4_K_M.gguf", runtime = rt)
model

gen <- rllm_generate(model, charToRaw(
    "The capital of France is Paris. The capital of Germany is"), n_new = 16L)
rawToChar(gen$raw)
```

A recorded run with SmolLM2-135M `Q4_K_M` (30 layers, grouped-query
attention, a `q5_0`-dominant quantization mix) used a 12-token prompt and 128
greedy decode tokens. The median of three complete generations was 40.2
tokens/s on CPU and 69.7 tokens/s on the RTX 5050 CUDA path after its one-time
weight upload.

`rllm_gguf_model()` reads the hyperparameters and borrows every weight's exact
read-only span. No weight is copied into the Rfmalloc work file. On CPU,
`rllm_forward()` points GGML tensors at those bytes while assembling RMSNorm,
RoPE, causal grouped-query attention and SwiGLU through Rggml's C-callables;
Rllm links no private GGML copy.

`rllm_kv_cache()` allocates one key and value slab per layer in plain R memory
or in fmalloc storage. The latter makes the cache file-backed, evictable and
open to a different cache codec without changing the graph. `rllm_generate()`
prefills that cache and then performs EOS-aware decoding one token at a time.

## Correctness discipline

The forward pass is pinned to a pure-R implementation of the same arithmetic:
about 1e-6 relative on the f32 twin of a real model and below 1e-4 on hermetic
MHA and GQA models, with separate causality probes. Incremental cache logits
must equal whole-batch logits at every position. Every codec decoder must be
bit-identical to GGML's type-traits `to_float`, the cross-validation that
caught the Q4_K bug in the earlier partial C port. At the byte boundary,
`rllm_decode(rllm_encode(x))` must always equal `x`.

The hermetic tests write their synthetic GGUF models at test time with
[Rgguf](../Rgguf)'s own writer; an opt-in smoke test
(`RLLM_TEST_GGUF=<path to a llama-arch GGUF>`) exercises real files.

## Compute follows residency

The CPU path borrows the GGUF mapping directly and uses GGML's
runtime-dispatched quantized kernels: AVX2 selected by CPUID on x86 and the
mandatory NEON baseline on aarch64, with no non-portable flags recorded in the
package. Dense products remain available to R's BLAS.

The optional CUDA path executes the same graph. Its model-owned context uploads
the codec-native weights once and retains them in device memory; subsequent
forward passes move only mutable inputs, cache state and logits through
Rggml's backend-neutral transfer interface. The host GGUF mapping remains the
authoritative storage, while device residency is an execution decision rather
than another model format. Vulkan also ships in Rggml, but Rllm does not yet
select it. The rig suite checks whole-batch and cached CUDA logits, plain and
fmalloc cache slabs, and cache handoff in both directions between CPU and CUDA.

Persistent allocation by itself is not the missing acceleration. A measured
GGML scheduler experiment reduced CUDA decode from 69.7 to 63.7 tokens/s, and
CUDA graph capture reached 62.6 because the attention extent and graph
properties change at every token. Both experiments were removed from the
default path. Stable graph shapes and a device-resident mutable KV cache are
the next execution questions; the storage and GGUF interfaces need no fork for
either.

## Install

From the [r-universe](https://sounkou-bioinfo.r-universe.dev):

```{r install, eval = FALSE}
install.packages("Rllm",
  repos = c("https://sounkou-bioinfo.r-universe.dev", getOption("repos")))
```

or via a GitHub subdir ref: `pak::pak("sounkou-bioinfo/Rfmalloc/packages/Rllm")`.
The CPU stack targets Linux, macOS and Windows; the optional CUDA build is
currently Unix-like. GPL (>= 2).
