---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(Rpgen)
```

# Rpgen

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/Rfmalloc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rfmalloc/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Rpgen is a native PLINK 2 `.pgen` reader for R, and the genomics I/O arm of the
[Rfmalloc](https://github.com/sounkou-bioinfo/Rfmalloc) stack. It vendors
PLINK 2's own `pgenlib` and exposes it through a **C-callable API** so that other
packages can read genotypes, dosages, phase and multiallelic records straight
from a `.pgen` in C, then pack them into Rfmalloc's file-backed genotype codecs
(`bed` at 2 bits/genotype, `dosage` at 1 byte, `haplotypes` at 1 bit) without a
dense double copy.

## Why it exists

The CRAN package [`pgenlibr`](https://cran.r-project.org/package=pgenlibr) also
vendors `pgenlib`, but it exposes only an R-level interface: it installs no
headers, declares no `LinkingTo`, and registers no `R_RegisterCCallable` entry
points. Nothing else can call its readers from C. Rpgen vendors the same library
and registers a C-callable surface (`Rpgen_open_info`, `Rpgen_read_hardcalls`,
`Rpgen_read_dosages`), which is the whole point: the Rfmalloc side reads `.pgen`
in C and writes into memory-mapped storage, with no round trip through an R
matrix.

## Provenance and license

`pgenlib` is **LGPL-3**, so Rpgen is **GPL-3** (LGPL-3 combines up to GPLv3, not
the GPL (>= 2) used elsewhere in the stack). The vendored tree is **generated**,
not hand-maintained: `tools/vendor-pgenlib/vendorpgen.R` reproduces it from a
pinned CRAN `pgenlibr` source tarball (the same "own our path" recipe the stack
uses for GGML), and CI guards it against drift
(`vendored-pgenlib-matches-recipe`). See `inst/COPYRIGHTS` for the full LGPL /
MIT / BSD attribution of `pgenlib`, `zstd`, `libdeflate` and `simde`.

## Usage

`rpgen_info()` opens a `.pgen` and reports its sample and variant counts,
straight from the header. These examples run against the small
`chr21_phase3_start.pgen` fixture that ships with the package:

```{r info}
pgen <- system.file("extdata", "chr21_phase3_start.pgen", package = "Rpgen")
rpgen_info(pgen)
```

The counts, and every genotype and dosage read below, are cross-checked against
`pgenlibr` reading the same file in the package's own tests (bit-exact, no
tolerance), so the native reader is pinned to the reference implementation.

`rpgen_dosage()` and `rpgen_bed()` read the genotypes in C and pack them
straight into Rfmalloc's genotype codecs, so the whole path from a `.pgen` on
disk to principal components never materializes a dense double genotype matrix:

```{r pca, message = FALSE}
library(Rfmalloc)
rt <- open_fmalloc(tempfile(fileext = ".bin"), mode = "scratch", size_gb = 1)

dose <- rpgen_dosage(pgen, runtime = rt)   # 1 byte/dosage, read from .pgen in C
c(payload_bytes = length(unclass(dose)),
  vs_double     = prod(dim(dose)) * 8 / length(unclass(dose)))

# Standardization (per-variant centre and scale, missing -> mean) is baked into
# the codec's decode, so PCA runs over the compressed tensor with center = FALSE.
std <- fmalloc_dosage_standardize(dose, runtime = rt)
pca <- fmalloc_pca(std, k = 5, center = FALSE)
round(pca$sdev, 2)
```

```{r pca_cleanup, include = FALSE}
cleanup_fmalloc(rt)
```

## Installation

From the [r-universe](https://sounkou-bioinfo.r-universe.dev):

```{r eval = FALSE}
install.packages("Rpgen",
  repos = c("https://sounkou-bioinfo.r-universe.dev", getOption("repos")))
```

or a GitHub subdir ref:

```{r eval = FALSE}
pak::pak("sounkou-bioinfo/Rfmalloc/packages/Rpgen")
```

Rpgen depends on `Rfmalloc`. It builds on Linux, macOS and Windows (Rtools) and
is checked on ARM (Apple Silicon and Linux aarch64) in CI.
