---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(Rfmalloc)
library(RfmallocStatgen)
```

# RfmallocStatgen

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/Rfmalloc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rfmalloc/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

RfmallocStatgen is the statistical-genetics layer of the
[Rfmalloc](https://github.com/sounkou-bioinfo/Rfmalloc) stack: it runs
genome-scale methods over fmalloc-backed genotype tensors (2-bit `bed`,
1-byte `dosage`), closing over the analysis stack popularized by
[bigsnpr](https://privefl.github.io/bigsnpr/)/[bigstatsr](https://privefl.github.io/bigstatsr/)
on denser storage with native readers
([Rpgen](https://github.com/sounkou-bioinfo/Rfmalloc): PGEN, BED, BGEN,
VCF/BCF, GEN, HAPS, and legacy dosage), and
adding the haplotype methods bigsnpr cannot express.

## Current closure and open programme

The implemented surface separates methods by the storage they actually need:

- genotype tensors: `statgen_gwas_lin()`, streamed randomized-SVD
  `statgen_pca()`, and `statgen_snp_cor()`;
- banded LD stores: `statgen_ldpred2_inf()`, `statgen_ldpred2()`, and
  `statgen_ldpred2_auto()`;
- signal matrices: `statgen_coloc_bf()`, with the all-pairs coupling expressed
  as one backend-dispatched matrix multiply.

The package tests these against direct dense references and, when installed,
the corresponding bigsnpr/bigstatsr routines. PCA is the explicit exception to
the clean-room rule: its randomized-SVD workhorse is pinned from GPL-3 PCAone
with deterministic dependency trimming and an R-console adaptation.

The active directions are logistic regression, clumping and pruning, GRM,
autoSVD and projection, lassosum2, LD score regression, split-LD, and direct
locus-row access for Li and Stephens or kalis-class HMMs. GPU composition stays
behind the same backend boundary; the current WSL rig exposes its RTX 5050 only
through CUDA, not Vulkan. See `ROADMAP.md` for the same list without the
rendered example.

## Usage

`statgen_gwas_lin()` regresses a phenotype on every variant in a `bed`/`dosage`
fmalloc genotype tensor, fused across variants (residualize once on the
covariate design, then reduce per variant), not a loop of per-variant `lm()`
calls. This toy example plants one causal variant and checks the fitted
effect against `lm()` directly:

```{r gwas}
rt <- open_fmalloc(tempfile(fileext = ".bin"), mode = "scratch", size_gb = 0.1)

set.seed(1)
n <- 300L
m <- 20L
g <- matrix(sample(0:2, n * m, replace = TRUE, prob = c(0.25, 0.5, 0.25)),
            nrow = n, ncol = m)
storage.mode(g) <- "integer"
covar <- cbind(age = rnorm(n, 50, 10), sex = rbinom(n, 1, 0.5))
causal <- 5L
y <- 0.8 * g[, causal] + 0.05 * covar[, "age"] + 0.3 * covar[, "sex"] + rnorm(n)

tn <- fmalloc_bed(g, runtime = rt)
gwas <- statgen_gwas_lin(tn, y, covar = covar)

gwas[causal, ]

# Cross-checked against a direct lm() fit at the causal variant.
fit <- lm(y ~ covar[, "age"] + covar[, "sex"] + g[, causal])
coef(summary(fit))[4L, ]

cleanup_fmalloc(rt)
```

The `beta`/`se`/`t`/`p` columns match `lm()`'s coefficient row for every
variant, not just the causal one; see `inst/tinytest/test_gwas_lin.R`.

## Provenance and license

RfmallocStatgen is GPL-3, matching bigsnpr and bigstatsr (both GPL-3): no
code from either is vendored, only their published algorithms and their
output as a numerical oracle for this package's own tests.
